---
title: "Storm Data - Code Book"
author: "Hugo van den Berg"
date: "August 21, 2016"
output: 
    github_document:
        toc: true
        toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = file.path(getwd(), '..'))
library(readr)
library(xtable)
library(magrittr)
library(tidyr)
```

## Introduction

This document describes the variables of the Storm Database, including
possible ranges.
As the original data seem to be poorly documented this code book describes
both the [original data](#original-data) as well as the 
[preprocessed data](#preprocessed-data).
Although with the preprocessed variables is described on what original 
variable(s) it is based, the exact transformation can be found in the
[main report](../reports/StormData.md)

Since it is an R Markdown based codebook, this actually generates the
transformation files for factor variables used in the main report.

## Original data

### STATE__

`STATE__` is a numeric variable, ranging from [x1, x2].
Its meaning is unclear, it appears to be a numeric representation of the state
abbreviation (see [STATE](#state)).

### BGN_DATE

`BGN_DATE` is a date variable ranging from 1950-01-03 until 2011-11-30.
It describes the start date of a weather event in the local timezone (see
[TIME_ZONE](#time_zone)).
Note that the original file includes a time in this variable.
However this is 00:00:00 throughout the file and can be ignored, and the value
from the [BGN_TIME](#bgn_time) should be used instead.

### BGN_TIME

`BGN_TIME` is a time variable containing the local time of day for the start of
a weather event.
It should be used together with [BGN_DATE](#bgn_date) and
[TIME_ZONE](#time_zone).

### TIME_ZONE

`TIME_ZONE` is a three letter code describing the timezone in which the event
occurred.

Timezones are inherently difficult to work with, and three letter POSIX codes
cannot be uniquely mapped to the Olson timezone names R uses.
Therefore the codes have been mapped manually to a suitable Olson timezone:

```{r tz, results='asis'}
source(file.path('doc', 'TZ.R'))
write_csv(TZ, file.path('data', 'TZ.csv'))
xTZ <- xtable(TZ)
names(xTZ) <- c("Time Zone", "State Code", "Olson/IANA Timezone")
print(xTZ, type = "html", include.rownames = FALSE,
      html.table.attributes = "")
```

*This table is available in [TZ.csv](../data/TZ.csv)*

### COUNTY

`COUNTY` is a numeric variable ranging from 0 to 873, describing the county in
which the weather event started.

### COUNTYNAME

`COUNTYNAME` is a character variable containing the county or counties in which
the event occurred.
It contains both references to county names and to county codes prepended with
the [state code](#state).

### STATE

`STATE` is a two-letter character variable containing an abbreviated state
name.
It contains 79 unique values, 50 of which can be related to the built in
`state.abb` variable.

Using the combination of state code and [countyname](#countyname) the meaning 
of the other 22 values was inferred.
Below is the ammended list of the used codes with their full description.

```{r state, results='asis'}
source(file.path('doc', 'STATE.R'))
write_csv(STATE, file.path('data', 'STATE.csv'))
xSTATE <- xtable(STATE)
names(xSTATE) <- c("State Code", "Region Name")
print(xSTATE, type = "html", include.rownames = FALSE,
      html.table.attributes = "")
```

*This table is available in [STATE.csv](../data/STATE.csv)*

### EVTYPE

`EVTYPE` is character variable containing the type of weather event that
occurred.
It contains 977 unique values (890 unique values ignoring upper/lower case),
ranging from "?" to "WND".

A large number of misspelled eventtypes was present.
The mapping below has been used to create uniform event types.

```{r evtype, results='asis'}
source(file.path('doc', 'EVTYPE.R'))
write_csv(EVTYPE, file.path('data', 'EVTYPE.csv'))
xEVTYPE <- xtable(EVTYPE)
names(xEVTYPE) <- c("Event type (original)", "Event type (unified)")
print(xEVTYPE, type = "html", include.rownames = FALSE,
      html.table.attributes = "")
```

*This table is available in [EVTYPE.csv](../data/EVTYPE.csv)*

### BGN_RANGE

`BGN_RANGE` is a numeric variable of which the meaning is unknown.
It contains 272 unique values ranging from 0 to 3749.

### BGN_AZI

`BGN_AZI` is a character variable describing the azimuth of the trajectory of
the weather event.

### BGN_LOCATI

`BGN_LOCATI` is a character variable giving a brief description of the starting
location of the weather event.

### END_DATE

`END_DATE` is a date variable ranging from 1986-04-10 until 2011-11-30, with
243411 missing values.
See also the notes on [BGN_DATE](#bgn_date).

### END_TIME

`END_TIME` is a time variable describing the end time of the weather event.

### COUNTY_END

`COUNTY_END` is the county code of the county in which the event ended.
In this dataset it is all 0.

### COUNTYENDN

`COUNTYENDN` is the name of the county in which the weather event ended, with
only missing values.

### END_RANGE

`END_RANGE` is a numeric variable of which the meaning is unknown.
It contains 266 unique values ranging from 0 to 925.

### END_AZI

`END_AZI` is the azimuth of the trajectory at the moment the weather event
ended.

### END_LOCATI

`END_LOCATI` is a character variable describing the location at which the event
ended.

### LENGTH

`LENGTH` is a numeric variable describing the length, in miles, of the path of
a tornado, and ranges from 0 to 2315.

### WIDTH

`WIDTH` is a numeric variable describing the width, in yards, of the path of a
tornado, and ranges from 0 to 4400.

### F

`F` is a numeric variable ranging from 0 to 5, with 843563 missing values.
It describes the force of a tornado using the Enhanced Fujita scale.

### MAG

`MAG` is a numeric variable ranging from 0 to 22000.
It is unknown what this variable describes, although it seems to describe the
magnitude of an event, with unknown units.

### FATALITIES

`FATALITIES` is a numeric variable ranging from 0 to 583, that denotes the
number of fatal victims caused directly or indirectly by the event.

### INJURIES

`INJURIES` is a numeric variable ranging from 0 to 1700 which denotes the
number of injured victims caused directly or indirectly by the event.

### PROPDMG

`PROPDMG` is a numeric variable ranging from 0 to 5000, and describes the total
damages in US dollars to properties caused by the weather event.
Note that the magnitude of the variable can be altered by the
[PROPDMGEXP](#propdmgexp) variable.

### PROPDMGEXP

`PROPDMGEXP` is a character variable that describes the magnitude of the
[PROPDMG](#propdmg) variable.
According to the [documentation][1] (page 12) it should only contain values "",
"K", "M", "B", for dollars, thousands of dollars, millions of dollars, and
billions of dollars, respectively.
It does however contain a number of other characters which are undocumented.

### CROPDMG

`CROPDMG` is a numeric variable ranging from 0 to 990, describing the damages
to crops in US dollars.
The magnitude of the value can be altered by the [CROPDMGEXP](#cropdmgexp)
variable.

### CROPDMGEXP

`CROPDMGEXP` is a character variable that describes the magnitude of the
[CROPDMG](#cropdmg) variable.
The same notes apply to this variable as to the [PROPDMGEXP](#propdmgexp)
variable.

### WFO

`WFO` is a character variable that denotes the responsible Weather Forecasting
Office.

### STATEOFFIC

`STATEOFFIC` is a character variable that denotes a state and region name.
Although the exact meaning is unclear it is probable that it signifies the
regional office within the state that was responsible for the report.

### ZONENAMES

`ZONENAMES` is a character variable describing the zonenames along the
trajectory of the weather event.

### LATITUDE

`LATITUDE` is a numeric variable ranging from 0 to 9706, with 47 missing
values.
It describes the latitude in hundreds of degrees north of the equator at which
the event started.

### LONGITUDE

`LONGITUDE` is a numeric variable ranging from -14451 to 17124, with 0 missing
values.
It describes the longitude in hundreds of degrees at which the event started.
As most of the values are centered around 10000, or 100 degrees, it is probable
that it describes the longitude in degrees *west* of the 0 meridian.

### LATITUDE_E

`LATITUDE_E` is a numeric variable ranging from 0 to 9706, with 40 missing
values.
It describes the latitude in hundreds of degrees north of the equator at which
the event started.

### LONGITUDE_

`LONGITUDE_` is a numeric variable ranging from -14455 to 106220, with 0
missing values.
It describes the longitude in hundreds of degrees at which the event started.
As with the [starting longitude](#longitude), most of the values are centered
around 10000, or 100 degrees, it is probable that it describes the longitude in
degrees *west* of the 0 meridian.

### REMARKS

`REMARKS` is a character variable containing remarks on the event, as well as
the event and episode narratives.

### REFNUM

`REFNUM` is a numeric variable ranging from 1 to 902300, with the recordnumber
within the dataset.


## Preprocessed data

### event.type.major

```{r etma}
EVTYPE %<>% separate(to, c("event.type.major", "event.type.minor"),
                     sep = "/", extra = "merge", fill = "right", 
                     remove = FALSE)
```
`event.type.major` is a factor variable containing uniform event types,
based on the original [EVTYPE](#evtype) and the listed transformations.
They contain only the part before the first "/" in the original variable.
`event.type.major` contains `r length(unique(EVTYPE$event.type.major))`
unique values, ranging from "`r min(EVTYPE$event.type.major, na.rm = T)`" to 
"`r max(EVTYPE$event.type.major, na.rm = T)`".

### fatalities

`fatalities` describes the number of people perished as either a direct or
indirect effect of the weather event.
It is a numeric variable ranging from 0 to 583.

### injuries

`injuries` describes the number of people injured as either a direct or
indirect effect of the weather event.
It is a numeric variable ranging from 0 to 1700.

### property.damages

`property.damages` expresses the damages to properties in US dollars.
It is a numeric variable ranging from $0.00 to $115 billion.

### crop.damages

`crop.damages` expresses the damages to crops in US dollars.
It is a numeric variable ranging from $0.00 to $5 billion.

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
