---
title: "Storm Data - Code Book"
author: "Hugo van den Berg"
date: "August 21, 2016"
output: 
    html_document:
        keep_md: true
        toc: true
        toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = file.path(getwd(), '..'))
library(readr)
library(xtable)
```

## Introduction

This document describes the variables of the Storm Database, including
possible ranges.
As the original data seem to be poorly documented this code book describes
both the [original data](#original-data) as well as the 
[preprocessed data](#preprocessed-data).
Although with the preprocessed variables is described on what original 
variable(s) it is based, the exact transformation can be found in the
[main report](../reports/StormData.md)

Since it is an R Markdown based codebook, this actually generates the
transformation files for factor variables used in the main report.

## Original data

### STATE__

`STATE__` is a numeric variable, ranging from [x1, x2].
Its meaning is unclear, it appears to be a numeric representation of the state
abbreviation (see [STATE](#state)).

### BGN_DATE

`BGN_DATE` is a date variable ranging from 1950-01-03 until 2011-11-30.
It describes the start date of a weather event in the local timezone (see
[TIME_ZONE](#time_zone)).
Note that the original file includes a time in this variable.
However this is 00:00:00 throughout the file and can be ignored, and the value
from the [BGN_TIME](#bgn_time) should be used instead.

### BGN_TIME

`BGN_TIME` is time variable containing the local time of day for the start of a
weather event.
It should be used together with [BGN_DATE](#bgn_date) and
[TIME_ZONE](#time_zone).

### TIME_ZONE

`TIME_ZONE` is a three letter code describing the timezone in which the event
occurred.

### COUNTY

`COUNTY` is a numeric variable ranging from 0 to 873, describing the county in
which the weather event started.

### COUNTYNAME

### STATE

```{r, results='asis'}
STATE <- data.frame(
    from = c(state.abb, "GU"),
    to = c(state.name, "Guam")
    )
write_csv(STATE, file.path('data', 'STATE.csv'))
names(STATE) <- c("State Code", "State Name")
print(xtable(STATE), type = "html", include.rownames = FALSE,
      html.table.attributes = "")
```

### EVTYPE

### BGN_RANGE

### BGN_AZI

### BGN_LOCATI

### END_DATE

### END_TIME

### COUNTY_END

### COUNTYENDN

### END_RANGE

### END_AZI

### END_LOCATI

### LENGTH

### WIDTH

### F

### MAG

### FATALITIES

### INJURIES

### PROPDMG

### PROPDMGEXP

### CROPDMG

### CROPDMGEXP

### WFO

### STATEOFFIC

### ZONENAMES

### LATITUDE

### LONGITUDE

### LATITUDE_E

### LONGITUDE_

### REMARKS

### REFNUM

## Preprocessed data

### event.type

`event.type` is a factor variable containing the uniform event types, based on
the original [EVTYPE](#evtype) and the listed transformations.

### state

### begin.date

### begin.county

### begin.latitude

### begin.longitude

### end.date

### end.county

### end.latitude

### end.longitude

### fatalities

### injuries

### property.damages

### crop.damages
